---
title: "Analysis of results AST set"
output: html_document
date: "2024-10-07"
---

```{r setup, include = FALSE}
# Set working directory
knitr::opts_knit$set(root.dir = '~/Documents/Projects/resistance_is_futile/amr-2024-team-prasad-feldgarden')
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

# Introduction

This notebook compares the results obtained using the `scanner.R` and `DGW.py` scripts. The objective is to assess whether both tools are yielding the same results, and if that is not the case, identify areas of opportunity to improve the scripts. In addition, the outputs from the scripts will be compared to the mutations found for those genomes in MicroBIGG-E.

# Setup

A total of 928 genome accession IDs for carbapenem resistant and sensitive *Pseudomonas aeruginosa* and *Klebsiella pneumoniae* were obtained from Antimicrobial susceptibility test (AST) data as follows:

*Pseudomonas aeruginosa* 
359 isolates

  - 292 with resistant phenotype
  - 67 with sensitive phenotype

*Klebsiella pneumoniae*
569 isolates

  - 297 with resistant phenotype
  - 272 with sensitive phenotype
  
The accession numbers for these genomes are found in the project repository [here](https://github.com/NCBI-Codeathons/amr-2024-team-prasad-feldgarden/tree/main/data/dev-examples/resis_and_sens_Kp_and_Pa).

## MicroBIGG-E subset

A subset of the data available in MicroBIGG-E for the selected isolates was obtained from BigQuery and stored in the repository [here](https://github.com/NCBI-Codeathons/amr-2024-team-prasad-feldgarden/blob/main/results/microbigge_subset.csv). 

## DGW results: `scanner.R` and `DGW.py`

The scripts were ran as described [here](https://github.com/NCBI-Codeathons/amr-2024-team-prasad-feldgarden/tree/main/data/resis_and_sens_Kp_and_Pa).

## Combining output files

Individual output files for each genome screened were combined using the `results_combined.R` script ([here](https://github.com/NCBI-Codeathons/amr-2024-team-prasad-feldgarden/blob/3e9bb4e9dc2196f7f8f5ad2c5055133ebd35a87d/results/scanner.R_AST/results_combined.R)).

The combined output files from each script (`DGW.py` and `scanner.R`) were combined and modeled running the `dgw_combined_outputs.sql` model ([here](https://github.com/NCBI-Codeathons/amr-2024-team-prasad-feldgarden/blob/main/results/script_comparison/dgw_combined_outputs.sql)) in `DuckDB`. 

# Analysis of results

```{r}
# Install packges
library(tidyverse)
library(dplyr)


# Import data

## Microbigg-E subset
microbigge <-  read_csv("results/microbigge_subset_ast.csv")

## Combined and modeled output from both scanner.R and DGW.py scripts
#dgw_combined_outputs_0 <- read_csv("results/script_comparison/dgw_combined_outputs.csv")
dgw_combined_outputs <- read_csv("results/script_comparison/dgw_combined_outputs_ast.csv")

```

Let's first prepare the data to establish the comparisons/analysis.


### How many lesions are detected by each script?
  
  - `DWG.py` identified 522 lesions
  - `scanner.R` identified 506 lesions
  
```{r}
dgw_combined_outputs %>% 
  summarize(mutations_p = n_distinct(p_mutation_id, na.rm = TRUE),
            mutations_r = n_distinct(r_mutation_id, na.rm = TRUE)
  )
```
  
### How many lesions are found by both scripts? 

  - `DWG.py` and `scanner.R` identified the lesions in the same 507 genomes
    - 414 of those are the same lesions
  
```{r}
dgw_combined_outputs %>% 
  filter(!is.na(p_mutation_id) & !is.na(r_mutation_id)) %>% 
  summarize(in_both = n_distinct(record_id, na.rm = TRUE),
            same = n_distinct(case_when(p_mutation_id == r_mutation_id ~ record_id), na.rm = TRUE)
  )
```

### Which records have mutations identified by both scripts, but with different lesions identified by each script? 

- There were 93 records where a mutation was identified by both `DGW.py` and `scanner.R`, but the lesions identified by each script were different
  - Of these, 3 records also reported mutations in Microbigg-E, which were not the same as the ones reported by the DGW scripts
  
```{r}
DT::datatable(dgw_combined_outputs %>% 
  filter(!is.na(p_mutation_id)  & !is.na(r_mutation_id) & p_mutation_id != r_mutation_id ) %>% 
  select(asm_acc, p_element_symbol, r_element_symbol, mb_element_symbol, p_lesion, r_lesion, mb_lesion)
)
```
    
    
### How many lesions are detected only by `DGW.py`?

- `DGW.py` identified 19 records that `scanner.R` did not identify.
  - One of those records was also present in Microbigg-E but reporting a different lesion

```{r}
DT::datatable(dgw_combined_outputs %>% 
  filter(!is.na(p_mutation_id) & is.na(r_mutation_id)) %>% 
  select(asm_acc, p_element_symbol, mb_element_symbol, p_lesion, mb_lesion)
  )

```

### How many lesions are detected only by `scaner.R`? 

- `scanner.R` identified 3 records that `DGW.py` did not identify
  - None of those mutations were present in Microbigg-E

```{r}
DT::datatable(dgw_combined_outputs %>% 
  filter(!is.na(r_mutation_id) & is.na(p_mutation_id)) %>% 
  select(asm_acc, r_element_symbol, mb_element_symbol, r_lesion, mb_lesion)
  )

```

## How many of the lesions detected by each script are in Microbigg-E?

- Of the 193 lesions found in Microbigg-E
  - `DGW.py` identified 134
  - `scanner.R` identified 133 (also identified by `DGW.py`)

```{r}
DT::datatable(dgw_combined_outputs %>% 
  summarize(in_microbigge = n_distinct(case_when(!is.na(mb_element_symbol) ~ record_id), na.rm = TRUE),
            mb_p = n_distinct(case_when(!is.na(mb_element_symbol) & !is.na(p_element_symbol) ~ record_id), na.rm = TRUE),
            mb_r = n_distinct(case_when(!is.na(mb_element_symbol) & !is.na(r_element_symbol) ~ record_id), na.rm = TRUE),
            mb_combined = n_distinct(case_when(!is.na(mb_element_symbol) & (!is.na(p_element_symbol) | !is.na(r_element_symbol)) ~ record_id), na.rm = TRUE),
            mb_in_both = n_distinct(case_when(!is.na(mb_element_symbol) & !is.na(p_element_symbol) & !is.na(r_element_symbol) ~ record_id), na.rm = TRUE)
  )
)
```

## How many of the lesions detected by each script are NOT in Microbigg-E? (i.e. new mutations)

- The DGW scripts identified 395 lesions that were not present in Microbigg-E
  - 359 of which were found in resistant isolates
    - `DGW.py` identified 356 new lesions
    - `scanner.R` identified 342
    - 339 of the new lesions were identified by both scripts
  
```{r}
DT::datatable(dgw_combined_outputs %>% 
  filter(!grepl("sensitive", cohort)) %>% 
  summarize(new_p = n_distinct(case_when(is.na(mb_element_symbol) & !is.na(p_element_symbol) ~ record_id), na.rm = TRUE),
            new_r = n_distinct(case_when(is.na(mb_element_symbol) & !is.na(r_element_symbol) ~ record_id), na.rm = TRUE),
            new_combined = n_distinct(case_when(is.na(mb_element_symbol) & (!is.na(p_element_symbol) | !is.na(r_element_symbol)) ~ record_id), na.rm = TRUE),
            new_in_both = n_distinct(case_when(is.na(mb_element_symbol) & !is.na(p_element_symbol) & !is.na(r_element_symbol) ~ record_id), na.rm = TRUE)
  )
)
```

```{r, echo = FALSE}
union(dgw_combined_outputs %>% 
  filter(!grepl("sensitive", cohort) & !is.na(p_element_symbol)) %>% 
  mutate(result = case_when(is.na(mb_element_symbol) ~ "New",
                            !is.na(mb_element_symbol) ~ "In Microbigg-E"
                            )) %>% 
  group_by(result) %>% 
  summarize(script = "DWG.py",
            mutations = n_distinct(record_id))
  ,
dgw_combined_outputs %>% 
  filter(!grepl("sensitive", cohort) & !is.na(p_element_symbol)) %>% 
  mutate(result = case_when(is.na(mb_element_symbol) ~ "New",
                            !is.na(mb_element_symbol) ~ "In Microbigg-E"
                            )) %>% 
  group_by(result) %>% 
  summarize(script = "scanner.R",
            mutations = n_distinct(record_id))
) %>% 
  ggplot(., aes(fill=result, y=mutations, x=script)) + 
    geom_bar(position="stack", stat="identity", width=0.7) +
    scale_fill_manual(name="",values=c('#68a1b3', '#a65005')) +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_blank(),
      panel.background = element_blank(),
      text = element_text(size = 16)) +
    ylab("# of Mutations") + xlab("Script") +
    scale_x_discrete(labels = function(x) stringr::str_wrap(x, width = 20))
  
```


### Which genes have different lesions identified by each script compared against the lesions in Microbigg-E? 

- There were 9 records were a mutation was found for the same gene, but the lesions reported by the DGW scripts are different from the lesions reported in Microbigg-E
  - 6 of those records show the same lesion from `DGW.py` and `scanner.R`

```{r}
DT::datatable(dgw_combined_outputs %>% 
  filter((mb_element_symbol == p_element_symbol | mb_element_symbol == r_element_symbol) & (mb_lesion != p_lesion | mb_lesion != r_lesion)) %>% 
  select(asm_acc, 
         element_symbol = mb_element_symbol,
         mb_lesion, 
         p_lesion,
         r_lesion)
  )

```

### How many lesions in Microbigg-E are not detected by the DGW scripts? (show lesions)

- Out of the 193 mutations present in Microbigg-E, 60 were missed by both DGW scripts
  - `DGW.py` missed 59 Microbigg-E mutations
  - `scanner.R` missed 60 Microbigg-E mutations

```{r}
DT::datatable(dgw_combined_outputs %>% 
  summarize(in_microbigge = n_distinct(case_when(!is.na(mb_element_symbol) ~ record_id), na.rm = TRUE),
            missed_from_microbigge = n_distinct(case_when(!is.na(mb_element_symbol) & (is.na(p_element_symbol) | is.na(r_element_symbol)) ~ record_id), na.rm = TRUE),
            p_missed_from_microbigge = n_distinct(case_when(!is.na(mb_element_symbol) & is.na(p_element_symbol) ~ record_id), na.rm = TRUE),
            r_missed_from_microbigge = n_distinct(case_when(!is.na(mb_element_symbol) & is.na(r_element_symbol) ~ record_id), na.rm = TRUE)
  )
)

```
```{r, echo = FALSE}
union(dgw_combined_outputs %>% 
  filter(!is.na(mb_element_symbol)) %>% 
  mutate(result = case_when(is.na(p_mutation_id) ~ "Missed by DGW",
                            !is.na(p_mutation_id) ~ "Found mutations"
                            )) %>% 
  group_by(result) %>% 
  summarize(script = "DWG.py",
            mutations = n_distinct(record_id))
  ,
dgw_combined_outputs %>% 
  filter(!is.na(mb_element_symbol)) %>% 
  mutate(result = case_when(is.na(r_mutation_id) ~ "Missed by DGW",
                            !is.na(r_mutation_id) ~ "Found mutations"
                            )) %>% 
  group_by(result) %>% 
  summarize(script = "scanner.R",
            mutations = n_distinct(record_id))
) %>% 
  ggplot(., aes(fill=result, y=mutations, x=script)) + 
    geom_bar(position="stack", stat="identity", width=0.7) +
    scale_fill_manual(name="",values=c('#68a1b3', '#a65005')) +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_blank(),
      panel.background = element_blank(),
      text = element_text(size = 16)) +
    ylab("# of Mutations") + xlab("Script") +
    scale_x_discrete(labels = function(x) stringr::str_wrap(x, width = 20))
  
```

```{r}
DT::datatable(dgw_combined_outputs %>% 
               filter (!is.na(mb_element_symbol) & (is.na(p_element_symbol) | is.na(r_element_symbol))) %>% 
               select(asm_acc, mb_element_symbol, mb_lesion, p_lesion, r_lesion)
  )

```


## Mutations found in sensitive isolates 

In 29 "sensitive" genomes, the DGW scripts found 18 unique lesions got genes *cirA*, *mexR*, *OmpK35*, *OmpK36*, and *OprD*. All of these genes encode a porin protein, with the exception of *mexR*, which is a transcriptional repressor of efflux systems. While it was not expected to detect any mutations in the genomes with a sensitive phenotype, there have been previous reports in the literature of mutations to some of these porins present in strains susceptible to antibiotics, and there have been various discussions suggesting that expression of a carbapenemase in combination to a porin mutation are required to confer antibiotic resistance. A study investigating *K. pneumoniae* isolates from a chronic obstructive pulmonary disease (COPD) patient found that an isolate lacking expression of *OmpK35* and presenting a loss-of-function mutation in *OmpK36* (L359R), but without the expression of any lactaase or carbapenemase, was susceptible to ceftazidime-avibactam. Conversely, the study found that an isolate presenting the same loss of *OmpK35* and the mutated *OmpK36* in combination with the lactamase CTX-M-71 was resistant to ceftazidime-avibactam (MIC 32 mg/L) ([Zhao et al., 2024](https://doi.org/10.1080/21505594.2024.2348251)). A similar conclusion can be made from a study looking at horizontal gene transfer of carbapenemases in *K. pneumoniae*, showing that strains with a truncated *OmpK35* (AA89STOP) and a GD insertion in *OmpK36* (134GD) had different susceptibility/resistance levels to various antibiotics depending on which carbapenemase was present; for example, the presence of the porin mutations resulted in an increase in ceftazidime MIC from 0.125 mg/l to 0.5 mg/L (compared to WT porins), and the addition of the carbapenemases KPC-109 or KPC-3 resulted in an increase in MIC from 256 mg/L (with WT porins) to 512 mg/L (with mutated porins) ([Di Pilato et al., 2023](https://doi.org/10.1016/j.ijantimicag.2023.107030)).


```{r}
# How many mutations found in each Sensitive genome
DT::datatable(dgw_combined_outputs %>% 
  filter(grepl("sensitive", cohort)) %>% 
  filter(!grepl("resistant", cohort)) %>% 
  #summarize(n_distinct(asm_acc))
  select(element_symbol = p_element_symbol,
         p_lesion,
         r_lesion,
         p_lesion_type,
         r_lesion_type) %>% 
  unique()
)

  
```

